---
title: "Analytic Q-Function for MVN with Missing Data"
format:
  html:
    toc: true
    number-sections: true
    code-fold: true
    html-math-method: katex
crossref:
  eq-prefix: "Eq."
---

## Setup

For multivariate normal data with missing values, partition the data into observed and missing components.

**Conditional distribution of missing given observed:**

$$
\mu_{m|o} = \mu_m + \Sigma_{mo} \Sigma_o^{-1} (Y_o - \mu_o)
$$ {#eq-cond-mean}

$$
\Sigma_{m|o} = \Sigma_m - \Sigma_{mo} \Sigma_o^{-1} \Sigma_{mo}^T
$$ {#eq-cond-var}

## Q-Function Definition

The Q-function is the expected complete-data log-likelihood, where the expectation is over the conditional distribution of missing data:

$$
Q(\hat{\theta} | \hat{\theta}) = E_{Y^m | Y^o, \hat{\theta}} \left[\log p(Y | \hat{\theta})\right]
$$ {#eq-Q-def}

## Derivation

### Step 1: Factor the Joint Density

$$
Q(\hat{\theta} | \hat{\theta}) = E_{Y^m | Y^o, \hat{\theta}} \left[\log p(Y^m | Y^o, \hat{\theta}) + \log p(Y^o | \hat{\theta})\right]
$$ {#eq-factor}

Since $\log p(Y^o | \hat{\theta})$ does not depend on $Y^m$, it passes through the expectation:

$$
= E_{Y^m | Y^o, \hat{\theta}} \left[\log p(Y^m | Y^o, \hat{\theta})\right] + \log p(Y^o | \hat{\theta})
$$ {#eq-factor2}

### Step 2: Conditional Log-Likelihood

The conditional distribution $Y^m | Y^o, \hat{\theta}$ is multivariate normal with mean $\mu_{m|o}$ and covariance $\Sigma_{m|o}$. Therefore:

$$
\log p(Y^m | Y^o, \hat{\theta}) = -\frac{q}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o}| - \frac{1}{2}(Y^m - \mu_{m|o})^T \Sigma_{m|o}^{-1}(Y^m - \mu_{m|o})
$$ {#eq-cond-loglik}

where $q$ is the dimension of the missing data.

### Step 3: Take Expectation

Taking the expectation over $Y^m | Y^o, \hat{\theta}$:

$$
E_{Y^m | Y^o, \hat{\theta}}\left[\log p(Y^m | Y^o, \hat{\theta})\right] = -\frac{q}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o}| - \frac{1}{2}E_{Y^m | Y^o, \hat{\theta}}\left[(Y^m - \mu_{m|o})^T \Sigma_{m|o}^{-1}(Y^m - \mu_{m|o})\right]
$$ {#eq-expect1}

### Step 4: Evaluate the Quadratic Form

Using the identity for quadratic forms of centered Gaussian vectors:

$$
E[X^T A X] = \text{tr}(A \cdot \text{Var}(X)) \quad \text{when } E[X] = 0
$$ {#eq-trace-identity}

Since $Y^m | Y^o, \hat{\theta} \sim N(\mu_{m|o}, \Sigma_{m|o})$, we have $Y^m - \mu_{m|o}$ centered with variance $\Sigma_{m|o}$:

$$
E\left[(Y^m - \mu_{m|o})^T \Sigma_{m|o}^{-1}(Y^m - \mu_{m|o})\right] = \text{tr}\left(\Sigma_{m|o}^{-1} \Sigma_{m|o}\right) = \text{tr}(I_q) = q
$$ {#eq-trace-result}

### Step 5: Combine Terms

Substituting back:

$$
E_{Y^m | Y^o, \hat{\theta}}\left[\log p(Y^m | Y^o, \hat{\theta})\right] = -\frac{q}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o}| - \frac{q}{2}
$$ {#eq-combine}

## Result (Single Observation)

The Q-function contribution from the missing data **for a single observation $i$** is:

$$
Q_i(\hat{\theta} | \hat{\theta})_{\text{mis}} = -\frac{q_i}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o,i}| - \frac{q_i}{2}
$$ {#eq-result}

Or more compactly:

$$
Q_i(\hat{\theta} | \hat{\theta})_{\text{mis}} = -\frac{q_i}{2}\left(1 + \log(2\pi)\right) - \frac{1}{2}\log|\Sigma_{m|o,i}|
$$ {#eq-result-alt}

where:

- $q_i$ = number of missing variables for observation $i$
- $\Sigma_{m|o,i}$ = conditional covariance matrix for observation $i$'s missing pattern

## Aggregation Across Observations

For monotone missingness, observations fall into groups with the same missing pattern. Let:

- $K$ = number of distinct missing patterns
- $n_k$ = number of observations with pattern $k$
- $q_k$ = number of missing variables in pattern $k$
- $\Sigma_{m|o,k}$ = conditional covariance for pattern $k$ (same for all obs in group $k$)

The **total Q-function contribution from missing data** is:

$$
Q(\hat{\theta} | \hat{\theta})_{\text{mis}} = \sum_{k=1}^{K} n_k \left[ -\frac{q_k}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o,k}| - \frac{q_k}{2} \right]
$$ {#eq-aggregate}

This simplifies to:

$$
Q(\hat{\theta} | \hat{\theta})_{\text{mis}} = -\frac{1}{2}\sum_{k=1}^{K} n_k \left[ q_k(1 + \log(2\pi)) + \log|\Sigma_{m|o,k}| \right]
$$ {#eq-aggregate-simple}

## Interpretation

The Q-function contribution from missing data depends on:

1. **$q_k$**: The dimension of missing data in each pattern
   - Comes from $\text{tr}(\Sigma_{m|o}^{-1}\Sigma_{m|o}) = q_k$

2. **$\log|\Sigma_{m|o,k}|$**: The log-determinant of conditional covariance for each pattern
   - Captures uncertainty in the conditional distribution

3. **$n_k$**: The count of observations in each missing pattern
   - Weights each pattern's contribution

## Connection to Entropy

This result is the negative differential entropy of the conditional distribution:

$$
H(Y^m | Y^o) = \frac{q}{2}\left(1 + \log(2\pi)\right) + \frac{1}{2}\log|\Sigma_{m|o}|
$$ {#eq-entropy}

So:

$$
Q_i(\hat{\theta} | \hat{\theta})_{\text{mis}} = -H(Y_i^m | Y_i^o)
$$ {#eq-neg-entropy}

## Numerical Verification via Gauss-Hermite Quadrature

We verify the analytic formula using Gauss-Hermite quadrature, which provides machine-precision integration for Gaussian expectations.

**Simplified test case:**

- $p = 4$ dimensional MVN
- Single missing pattern: observe $(Y_1, Y_2)$, miss $(Y_3, Y_4)$
- $q = 2$ missing dimensions

```{r}
#| label: setup
#| message: false

library(MASS)
library(statmod)  # For gauss.quad

set.seed(42)
```

### Define Parameters

```{r}
#| label: params

# 4-dimensional MVN
p <- 4
mu <- c(1, 2, 3, 4)
Sigma <- toeplitz(0.6^(0:(p-1)))

cat("Mean vector:", mu, "\n")
cat("\nCovariance matrix:\n")
print(round(Sigma, 3))
```

### Partition into Observed/Missing

```{r}
#| label: partition

# Observe first 2, miss last 2
obs_idx <- 1:2
mis_idx <- 3:4
q <- length(mis_idx)

# Fixed observed values
Y_obs <- c(0.5, 1.5)

# Partition parameters
mu_o <- mu[obs_idx]
mu_m <- mu[mis_idx]
Sigma_oo <- Sigma[obs_idx, obs_idx]
Sigma_mm <- Sigma[mis_idx, mis_idx]
Sigma_mo <- Sigma[mis_idx, obs_idx]
Sigma_om <- Sigma[obs_idx, mis_idx]

# Conditional distribution parameters (Eqs. 1-2)
Sigma_oo_inv <- solve(Sigma_oo)
mu_m_given_o <- mu_m + Sigma_mo %*% Sigma_oo_inv %*% (Y_obs - mu_o)
Sigma_m_given_o <- Sigma_mm - Sigma_mo %*% Sigma_oo_inv %*% Sigma_om

cat("Conditional mean (mu_m|o):", as.vector(mu_m_given_o), "\n")
cat("\nConditional covariance (Sigma_m|o):\n")
print(round(Sigma_m_given_o, 4))
cat("\nlog|Sigma_m|o| =", log(det(Sigma_m_given_o)), "\n")
```

### Analytic Q-Function (@eq-result)

```{r}
#| label: analytic

Q_analytic <- -q/2 * log(2*pi) - 0.5 * log(det(Sigma_m_given_o)) - q/2

cat("Analytic Q_mis =", Q_analytic, "\n")
```

### Gauss-Hermite Quadrature

For a Gaussian integral with $Y^m \sim N(\mu_{m|o}, \Sigma_{m|o})$, we use the transformation via Cholesky decomposition: $Y^m = \mu_{m|o} + \sqrt{2} L^T z$ where $L^T L = \Sigma_{m|o}$.

```{r}
#| label: gauss-hermite

# Gauss-Hermite quadrature points and weights
n_quad <- 20
gh <- gauss.quad(n_quad, kind = "hermite")

# Cholesky decomposition
L <- chol(Sigma_m_given_o)

# Log conditional density function
log_cond_density <- function(Y_m) {
  diff <- Y_m - as.vector(mu_m_given_o)
  Sigma_inv <- solve(Sigma_m_given_o)
  -q/2 * log(2*pi) - 0.5 * log(det(Sigma_m_given_o)) -
    0.5 * sum(diff * (Sigma_inv %*% diff))
}

# Tensor product grid for 2D integration
grid <- expand.grid(z1 = gh$nodes, z2 = gh$nodes)

Q_numerical <- 0
for (i in 1:nrow(grid)) {
  z <- c(grid$z1[i], grid$z2[i])

  # Transform: Y_m = mu_m|o + sqrt(2) * L' * z
  Y_m <- as.vector(mu_m_given_o) + sqrt(2) * t(L) %*% z

  # Weight = w1 * w2 / pi
  w <- gh$weights[which(gh$nodes == grid$z1[i])] *
       gh$weights[which(gh$nodes == grid$z2[i])] / pi

  Q_numerical <- Q_numerical + w * log_cond_density(Y_m)
}

cat("Gauss-Hermite Q_mis =", Q_numerical, "\n")
```

### Comparison

```{r}
#| label: comparison

cat("\n=== Verification Results ===\n")
cat("Analytic Q_mis:      ", Q_analytic, "\n")
cat("Gauss-Hermite Q_mis: ", Q_numerical, "\n")
cat("Difference:          ", Q_analytic - Q_numerical, "\n")
```

## Summary

The Gauss-Hermite verification confirms the analytic formula to machine precision. The key result (@eq-result) is:

$$
Q_i(\hat{\theta} | \hat{\theta})_{\text{mis}} = -\frac{q}{2}\log(2\pi) - \frac{1}{2}\log|\Sigma_{m|o}| - \frac{q}{2}
$$

This can be computed directly from the conditional covariance $\Sigma_{m|o}$ without any Monte Carlo sampling.
